Transfer learning for echelle spectroscopy

### 2021-11-05 paper/paper1/ms.tex
Outline the entire paper in latex, with bullet points for themes

@@

- [ ]
- [ ] Underlying atomic and molecular data may be wrong or approximate or missing
- [ ] Underlying radiative transfer may be inaccurate or approximate (T-P profile / T-Tau)
- [ ] Time is a dimension: Star is changing
- [ ] Stellar surface inhomogeneities become important: Doppler Imaging and limb darkening
- [ ] We do not know how to accurately parameterize some dimensions
- [ ] High dynamic range is both a blessing and a curse:
- [ ] Feeding back locations of model imperfections becomes unwieldy for many lines
- [ ] Many lines essentially becomes a book-keeping and continuum assignment problem
- [ ] Computational problem
- [ ] Some lines have large line wings, blurring continuum and line
- [ ] Line blanketing has almost no continuum
- [ ] So in practice, tradeoffs between model flexibility and physical self-consistency
- [ ] Semi-empirical models may be a middle ground: informed from models, but revised with data
- [ ] The radiative transfer step is computationally expensive, so we want to avoid redoing that
- [ ] Instead focus on pre-computed models that have that wisdom baked-in
- [ ] We are essentially compressing pre-computed models into evaluable, interpretable models
- [ ] Transfer learning is one plausible approach: pre-train on pre-computed models, transfer to data
- [ ] Neural network frameworks make this possible with autodiff
- [ ] We use the autodiff and GPU acceleration, but not the neural architectures
- [ ] We do not use the neural architectures because we have good understand for how a spectrum is generated
- [ ] Instead of tuning weights of a neural network, we tune the atomic and molecular properties of a spectrum

\newpage

@@

TBD

\section{Conclusions}

More placeholder text...

@@

Here are some more details about autodiff
\end{document}

>>>>

\usepackage{outlines}
\usepackage{enumitem}

@@

\begin{itemize}
  \item Underlying atomic and molecular data may be wrong or approximate or missing
  \item Underlying radiative transfer may be inaccurate or approximate (T-P profile / T-Tau)
  \item Time is a dimension: Star is changing
  \item Stellar surface inhomogeneities become important: Doppler Imaging and limb darkening
  \item We do not know how to accurately parameterize some dimensions
\end{itemize}

\subsection{High dynamic range is both a blessing and a curse:}
\begin{itemize}
  \item Feeding back locations of model imperfections becomes unwieldy for many lines
  \item Many lines essentially becomes a book-keeping and continuum assignment problem
  \item Computational problem
  \item Some lines have large line wings, blurring continuum and line
  \item Line blanketing has almost no continuum
\end{itemize}

\subsection{This paper: towards semi-empirical models}
\begin{itemize}
  \item So in practice, tradeoffs between model flexibility and physical self-consistency
  \item Semi-empirical models may be a middle ground: informed from models, but revised with data
  \item The radiative transfer step is computationally expensive, so we want to avoid redoing that
  \item Instead focus on pre-computed models that have that wisdom baked-in
  \item We are essentially compressing pre-computed models into evaluable, interpretable models
  \item Transfer learning is one plausible approach: pre-train on pre-computed models, transfer to data
  \item Neural network frameworks make this possible with autodiff
  \item We use the autodiff and GPU acceleration, but not the neural architectures
  \item We do not use the neural architectures because we have good understand for how a spectrum is generated
  \item Instead of tuning weights of a neural network, we tune the atomic and molecular properties of a spectrum
\end{itemize}

@@

\subsection{Overall Architecture and Design Choices}
Here is an outline
\begin{outline}
  \1 Currently limit the bandwidth to a single high-bandwidth echelle spectrograph (plus buffer at edges)
  \2 This choice is principally computational at the moment, no fundamental limit
  \1 We fetch a high resolution pre-computed spectrum from PHOENIX
  \1 We use the exact native wavelength sampling, yielding N pixels
  \1 Tradeoff of working in log or linear flux
  \2 First normalize by a scalar since we do not intend to use absolute flux calibration
  \2 Log prevents negative fluxes
  \2 Trickier to do Pseudo-Voigt approximation as sum in linear space
\end{outline}

\subsection{Initialization}
\begin{outline}
  \1 Initialize the model with preprocessing steps:
  \2 Low-pass filter to remove narrow spectral lines
  \2 (Optional) Place known Hydrogen, Sodium lines with large wings and non-standard line shapes
  \2 Define a prominence threshold and find locations of peaks above that threshold
  \2 Derive coarse properties about those peaks: widths and amplitudes
  \1 We now have a coarse model that needs to be fine-tuned with autodiff
\end{outline}

\subsection{Continuum model}
\begin{outline}
  \1 How to describe the continuum:
  \2 Black body / Rayleigh Jeans plus perturbation
  \2 High order polynomial: Chebyshev? Somewhere between 3rd and 20th order?
  \2 Use cross validation?
  \1 GPs are also possible, but high Number of pixels--> need scalable GP
  \2 Celerite not implemented for PyTorch?
  \2 GPyTorch possible, but has certain restrictions on kernel and sampling
\end{outline}

\subsection{Lineshapes choice and tradeoffs}
\begin{outline}
  \1 Faced with a question: what lineshape to use
  \2 To First order the lines are Gaussian or Lorentzian
  \2 To Second order the lines are Voigt Profiles
  \2 In detail the lines are the result of radiative transfer that can smear the Voigt profile
  \2 In practice the high-res lines get convolved with instrumental profiles, so details don't matter too much
  \1 Computational tradeoff of Lorentzian/Gaussian versus Pseudo-Voigt versus Voigt
  \2 Additional Implementation challenge: Fadeeva function not implemented in PyTorch
  \1 We choose a pseudo-Voigt as a balance between computation and adequate accuracy
\end{outline}

\subsection{GPU and Autodiff specific considerations}
\begin{outline}
  \1 Total number of parameters
  \2 For the lines =  N-lines x 4 (lambda center, fwhm L, sigma G, Amplitude)
  \2 For the continuum = Nth order polynomial = N+1 parameters, or Gaussian Process = 2 hyperparameters
  \1 Each Tunable parameter gets `grad=True`
  \1 We assemble the product of all lines and all continuum sources, this is the computational bottleneck
  \1 This becomes a matrix of size N-lines x N-wavelength points
  \1 We take the element-wise product over the lines axis, contracting to 1 x N-wavelength points
  \1 For N-lines = X and N-wavelength-points = Y, we expect X*Y*k primative operations.
  \1 For N-lines = X and N-wavelength-points = Y, we expect X*Y*32 bits of memory => XX GB
  \1 NVIDIA GPUs can handle 8 - 40 GB of memory depending on the model
\end{outline}

\subsection{Minibatches}
\begin{outline}
  \1 In the parlance of Machine Learning, we have a big model and small data
  \1 Really our model is too big: most of the entries are zero, so we should use a sparse implementation
  \1 If the model is too big to fit in memory, we have to train in mini-batches
  \1 We hand-in some fraction f ~ 1/20th of the entire wavelength range as a mini-batch
  \1 The model is only evaluated at that 5\% of pixels,  $N_{lines} \times N_{pix} \cdot 0.05$)
  \1 Technically all lines are updated
  \1 In practice, only lines that are sufficiently informed by those pixels will get large updates
  \1 Causes some stochasticity--- Stochastic Gradient Descent "SGD"
  \1 Analogous but different from the notion of minibatch in ML: not a distinct axis, like $N_{images}$
  \1 We use random minibatches
\end{outline}

\subsection{Optimization and training}
\begin{outline}
  \1 Run the training:
  \1 Set the optimizer: Adam with a LR=0.004
  \1 Set the loss function: Chi-squared with no per-pixel uncertainty
  \1 Set the "mini-batch" fraction f~1-5%
  \1 Choice of epochs or steps $N_steps = 300-10000$
  \1 The training process sees 20 minibatches per epoch, sees $N_pixels$, with some repeats some misses
  \1 Monitor with Tensorboard: see the improvement
  \1 The training takes X minutes on an RTX2070 GPU with PyTorch v1.11.x, CUDA v11.x, and Intel X CPU
  \1 We save the model parameters, refer to them collectively as the pre-trained model
\end{outline}

\section{Cloning Performance}

\subsection{Results and scaling}

\begin{outline}
  \1 We compute the residual of native PHOENIX minus cloned model
  \1 We see X percent residuals at native resolution
  \1 We smooth the residual to HPF instrumental resolution and see Y percent residuals
  \1 The main residuals come from missing line opacity due to our prominence threshold:
  \1 Including smaller prominence lines means smaller residuals, but more lines and higher computational cost
\end{outline}

\subsection{Line data clustering and conspicuous flaws}
\begin{outline}
  \1 We see some lines that devolve into missing continuum opacity
  \1 Especially acute near broad line wings if you don't handle the non-standard line shapes
  \1 How do the lineshapes cluster in $\sigma_G$ vs $FWHM_L$?
  \1 How do the lineshapes cluster in Amplitude vs FWHM? Can some lines be rejected?
  \1 Does the pre-trained model round-trip effectively (model saved at end of training = model from loaded params)?
\end{outline}

\section{Refining the model from data with transfer learning}
\begin{outline}
  \1 Augment the pre-trained model with extrinsic parameters:
  \2 vsini and RV
  \1 How to re-sample into the data space?
  \1 Option 1: Convolve the Pseudo-Voigt with a Instrumenal Gaussian
  \2 The convolution of a Gaussian with a Pseudo-Voigt is simply a new Pseudo-Voigt!
  \2 Simply evaluate on a new coarse wavelength grid!  Saves computation!
  \1 Option 2: Numerically convolve output on a fine grid and sum pixels
  \2 Computationally wasteful
  \2 Requires generating full bandwidth spectrum at native resolution
  \1 Both options should give the same answer (right?)
  \2 Need to double-check commutivity of convolution and element-wise product...
  \1 Ideally you need wavelength-dependent resolution, not just a single stock resolving power
  \1 Need accurate per-pixel uncertainties
  \1 We have to apply some amount of regularization to the pre-trained model
  \2 Regularization tunes how much to trust the data versus the model in the semi-empirical model
  \2 No Regularization means likely to overfit noise in the data
  \2 Too much regularization resembles a static (fixed) PHOENIX model
\end{outline}

\section{Performance of transfer learning}
\begin{outline}
  \1 What is the residual level with bare PHOENIX (i.e. tuning continuum only, not lines)?
  \1 What is the residual level with no regularization (i.e. tune all lines and continuum)?
  \2 Should be near-zero except for missing lines
  \2  What should we do about missing lines?
  \1 What is the residual level with modest regularization?  What is the typical change to the line?
  \2 Plot of cloned FWHM versus FWHM transferred
  \2 The pseudo-voigt line properties can be analytically integrated to give an equivalent width
  \2 Plot of cloned EW (before) versus transfer EW (after)
\end{outline}

\section{Discussion}

\subsection{Questioning the assumptions of the method}
\begin{outline}
  \1 Is PseudoVoigt Adequate?
  \1 What performance level is adequate? (depends on your application)
  \1 What about PRV applications
  \1 What about the choice of minibatch size?
  \1 How would a sparse tensor change the computational performance?
  \1 Are we using the GPU effectively?  Are we memory bandwidth limited?
  \1 Should we update the line center positions?
\end{outline}

\subsection{Limitations}
\begin{outline}
  \1 What to do when continuum is absent even in at the native resolution of precomputed synthetic model?
  \1 Relevant to brown dwarfs and possibly M dwarfs molecular bands
  \1 Bandwidth limitations
  \1 Interplay with tellurics
\end{outline}

\section{Conclusions}
More placeholder text...

\

@@

Here are some more details about autodiff

% Table
\begin{deluxetable}{cc}
  \tablecaption{Notation used in this paper\label{table2}}
  \tablehead{
    \colhead{Symbol} & \nocolhead{Meaning}
  }
  \startdata
  $N_{pix}$ & Number of pixels in the precomputed spectrum \\
  $N_{lines}$ & Number of spectral lines \\
  $\sigma_G$ & Gaussian line profile scale and standard deviation \\
  $\gamma_L$ & Lorentzian line profile half width
  \enddata
\end{deluxetable}

\end{document}
